Dissertation Proposal

### 2023-11-07 proposal.tex
try to switch to natbib
clean up citations

\usepackage[switch]{lineno}
\usepackage{xcolor}
\renewcommand\thelinenumber{\textcolor{red}{\arabic{linenumber}}}
\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
\usepackage[style=authoryear]{biblatex}  % For managing references
\addbibresource{bibfile.bib}
\usepackage{hyperref}
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{tabularx}   %% For tables with growable column
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\usepackage{cleveref}   %% For smart ref
\usepackage{graphicx}   %% For inserting images
\graphicspath{ {./images/} }  %% configure the graphicx package
\usepackage{multirow}
\usepackage{multicol}
\usepackage{ragged2e}  %% For \raggedright
\usepackage{listings}  %% For typesetting code
\usepackage{enumitem}  %% For noitemsep and topsep
\lstdefinelanguage{smol} {
  alsoletter={ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz-!?*},
  morekeywords={defvar,deffun,lambda,set!,let,let*,letrec,begin,if,cond,else,and,or},
  morecomment=[l]{;},
  morestring=[b]",
}
\lstset{
  basicstyle=\small\ttfamily,
  breaklines=true,
  language=smol
}

@@

\end{itemize}
This semantic core can be seen in languages from ``object-oriented''
languages like C\# and Java, to ``scripting'' languages like
JavaScript, Python, and Ruby, to ``functional'' languages like the ML
and Lisp families. Of course, there are sometimes restrictions
(e.g., Java has restrictions on closures) and extensions (such as the
documented semantic oddities of JavaScript and Python \cite{bernhardtWat2012, guhaEssenceJavascript2010, politzPythonFullMonty2013, politzTestedSemanticsGetters2012}). Still, this semantic core bridges many syntaxes, and
understanding it helps when transferring knowledge from old languages
to new ones. In recognition of this deep commonality, in this paper we
choose to call this the \emph{Standard Model of Languages} (SMoL).

@@

In fact, these problems are not at all limited to students. This
specific looping problem even trips up industrial programmers. The
C\# language \emph{changed} to produce \lstinline|0| through \lstinline|9| \cite{lippertClosingLoopVariable2009}. It is now also the focus of a language design
change in Go \cite{chaseFixingLoopsGo2023} for a new kind of looping
construct \cite{coxSpecLessErrorprone2023}.
It continues to trip up programmers
(e.g., \cite{sharveyCreatingFunctionsLambdas2022})
despite being
documented as a ``gotcha'' in Python \cite{reitzHitchhikerGuidePython2016}.
% SK: But not everyone has read the document. I don't think the document is
% a Python official document. Also, this sentence sounds like stating that
% the cited Python programmers are stupid.

Most of the time, however, these misunderstandings do not lead to
language design changes. Nor are changes necessarily desirable: the
SMoL feature set has presumably evolved because it is convenient
for writing non-trivial programs (e.g., mutable structures) without
being too unwieldy (e.g., no dynamic scope). Furthermore, having a
good mental model of these features
is essential to understand ownership \cite{clarkeOwnershipTypesFlexible1998},
manage parallelism, and more. Thus, we
still need to train students and other programmers on the semantic
features \emph{and their interactions} in SMoL.

@@

SMoL's Lispy syntax is
an artifact of the course (\Cref{s:populations}) using Racket
\cite{friedmanEssentialsProgrammingLanguages2001, krishnamurthiProgrammingLanguagesApplication2007};
however, it also proved to be pedagogically valuable.
I have found the parentheses useful when discussing
scope in conjunction with local-binding features like \lstinline|let|.
This avoids the various confusing ``variable lifting'' semantics found
in languages like Python (\cite{politzPythonFullMonty2013}; see also
posts like \cite{froadieWhatScopeVariable2022}), where the actual defined range of a
variable is not apparent from the source code.

@@

Students are given a working implementation of SMoL, built
as a \lstinline|#lang| language inside Racket \cite{felleisenProgrammableProgrammingLanguage2018}. This language
provides only the defined syntax and semantics of SMoL, with no
(other) Racket features present.
(As in Racket, arguments evaluate left-to-right, to give stateful programs an
unambiguous semantics.) The implementation is available online:
\href{https://github.com/shriram/smol}{https://github.com/shriram/smol}.

@@

There is an extensive body of literature on tutoring systems
(\cite{vanlehnBehaviorTutoringSystems2006a} is a quality survey), and
indeed whole conferences are dedicated to them. I draw on this
literature. In particular, it is common in the literature to talk
about a ``two-loop'' architecture
\cite{vanlehnBehaviorTutoringSystems2006a} where the outer loop
iterates through ``tasks'' (i.e., educational activities) and the
inner loop iterates through UI events within a task. I follow the same
structure in my tutor (\Cref{s:tutor-uiux}).

Many tutoring systems focus on teaching programming (such as the
well-known and heavily studied LISP Tutor
\cite{andersonLISPTutor1985}), and in the process undoubtedly address
some program behavior misconceptions. My tutor differs in a notable
way: it does not try to teach programming per se. Instead, it assumes
a basic programming background and focuses entirely on program
behavior misconceptions and correcting them. I are not aware of a
tutoring system (in computer science) that has this specific design.

@@

The SMoL Tutor is firmly grounded in one technique from
cognitive and educational psychology. The fundamental problem is: how
do you fix a misconception? One approach is to only ``present the
right answer'', for fear that discussing the wrong conception might
actually reinforce it. Instead, there is a body of literature starting
from \cite{posnerTheoryConceptualChange1982} that presents a theory of conceptual
change, at whose heart is the \emph{refutation text}. A refutation
text tackles the misconception directly, discussing and providing a
refutation for the incorrect idea. Several studies \cite{schroederRefutationTextFacilitates2022} have shown
their effectiveness in multiple other domains.

@@

The problem with expert-generated lists is that they can be quite
incomplete. Education researchers have documented the
phenomenon of the \emph{expert blind spot} \cite{nathanExpertBlindSpot2001}: experts
simply do not conceive of many learner difficulties.
Thus, we need
methods to identify problems beyond what experts conceive.

@@

Finally, we are inspired by the significant body of education research
on \emph{concept inventories} \cite{hestenesForceConceptInventory1992} (with a growing
number for computer science, as a survey lists \cite{taylorComputerScienceConcept2014}). In
terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question has
one correct answer and several wrong ones. However, the wrong ones are
chosen with great care. Each one has been validated so that if a
student picks it, we can quite unambiguously determine \emph{what
  misconception the student has}. For instance, if the question is
``What is \lstinline|sqrt(4)|?'', then \lstinline|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \lstinline|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many
  settings. For instance, an educator can use them with clickers to
  get quick feedback from a class. If several students pick a
  specific wrong answer, the educator not only knows they are wrong,
  but also has a strong inkling of \emph{precisely what} misconception
  that group has and can address it directly. I expect our
  instruments to be useful in the same way.}

@@

Our main solution to the expert blind spot is to use the Quizius system
\cite{saarinenHarnessingWisdomClasses2019}. In contrast to the very
heavyweight process (involving a lot of expert time)
that is generally
used to create a concept inventory, Quizius uses a lightweight,
interactive approach to obtain fairly comparable data, which an expert
can then shape into a quality instrument.

@@

Quizius runs interactively during an assignment period.
At each point, it needs to determine which
previously authored program to show a student.
It can
either ``exploit'' a given program that already has responses
or ``explore'' a new
one. Quizius thus treats this as a multi-armed bandit problem
\cite{katehakisMultiArmedBanditProblem1987} and uses that to choose a program.

@@

Prior work using Quizius \cite{saarinenHarnessingWisdomClasses2019} finds that students do author programs
that the experts did not imagine. In our case, we seeded Quizius with
programs from prior papers (\Cref{s:background-misconceptions}), which gives the first
few students programs to respond to. However, we found
that Quizius significantly expanded the scope of our problems and
misconceptions. In our final instrument, most programs were directly
or indirectly inspired by the output of Quizius.

@@

Is our reasoning valid? I have applied standard techniques and
measurements for evaluating student responses. However, our
instruments still lack the validity of a proper concept
inventory. Their creation requires heavyweight processes (such as
Delphi methods \cite{goldmanSettingScopeConcept2010}) that require many hours of expert attention as
well as conversations with learners, and can hence be prohibitive in
cost. The Quizius method (\Cref{s:quizius}) was created precisely to
be an inexpensive method that provides a good proxy.

@@

Our idea of misinterpreters is related to mystery languages
\cite{diwanPLdetectiveSystemTeaching2004,
  pombrioTeachingProgrammingLanguages2017}. Both approaches use
evaluators that represent alternative semantics to the same syntax.
However, the two are complementary.
In mystery languages, instructors design the space of semantics with
pedagogic intent, and
students must create programs to explore that space. Misinterpreters,
in contrast, are driven by student input, while the programs are provided
by instructors. The two approaches also have different goals:
mystery languages focus on encouraging students to experiment with
languages; misinterpreters aim at capturing students' misconceptions.

@@

\printbibliography

>>>>

\usepackage[switch]{lineno}
\usepackage{xcolor}
\renewcommand\thelinenumber{\textcolor{red}{\arabic{linenumber}}}
\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
% \usepackage[style=authoryear]{biblatex}  % For managing references
\usepackage[round, square]{natbib}
\bibliographystyle{plainnat}
\usepackage{hyperref}
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{tabularx}   %% For tables with growable column
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\usepackage{cleveref}   %% For smart ref
\usepackage{graphicx}   %% For inserting images
\graphicspath{ {./images/} }  %% configure the graphicx package
\usepackage{multirow}
\usepackage{multicol}
\usepackage{ragged2e}  %% For \raggedright
\usepackage{listings}  %% For typesetting code
\usepackage{enumitem}  %% For noitemsep and topsep
\lstdefinelanguage{smol} {
  alsoletter={ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz-!?*},
  morekeywords={defvar,deffun,lambda,set!,let,let*,letrec,begin,if,cond,else,and,or},
  morecomment=[l]{;},
  morestring=[b]",
}
\lstset{
  basicstyle=\small\ttfamily,
  breaklines=true,
  language=smol
}

@@

\end{itemize}
This semantic core can be seen in languages from ``object-oriented''
languages like C\# and Java, to ``scripting'' languages like
JavaScript, Python, and Ruby, to ``functional'' languages like the ML
and Lisp families. Of course, there are sometimes restrictions (e.g.,
Java has restrictions on closures) and extensions (such as the
documented semantic oddities of JavaScript and Python
\citep{bernhardtWat2012, guhaEssenceJavascript2010,
politzPythonFullMonty2013, politzTestedSemanticsGetters2012}). Still,
this semantic core bridges many syntaxes, and understanding it helps
when transferring knowledge from old languages to new ones. In
recognition of this deep commonality, in this paper we choose to call
this the \emph{Standard Model of Languages} (SMoL).

@@

In fact, these problems are not at all limited to students. This
specific looping problem even trips up industrial programmers. The
C\# language \emph{changed} to produce \lstinline|0| through \lstinline|9| \citep{lippertClosingLoopVariable2009}. It is now also the focus of a language design
change in Go \citep{chaseFixingLoopsGo2023} for a new kind of looping
construct \citep{coxSpecLessErrorprone2023}.
It continues to trip up programmers
(e.g., \cite{sharveyCreatingFunctionsLambdas2022})
despite being
documented as a ``gotcha'' in Python \citep{reitzHitchhikerGuidePython2016}.
% SK: But not everyone has read the document. I don't think the document is
% a Python official document. Also, this sentence sounds like stating that
% the cited Python programmers are stupid.

Most of the time, however, these misunderstandings do not lead to
language design changes. Nor are changes necessarily desirable: the
SMoL feature set has presumably evolved because it is convenient
for writing non-trivial programs (e.g., mutable structures) without
being too unwieldy (e.g., no dynamic scope). Furthermore, having a
good mental model of these features
is essential to understand ownership \citep{clarkeOwnershipTypesFlexible1998},
manage parallelism, and more. Thus, we
still need to train students and other programmers on the semantic
features \emph{and their interactions} in SMoL.

@@

SMoL's Lispy syntax is
an artifact of the course (\Cref{s:populations}) using Racket
\citep{friedmanEssentialsProgrammingLanguages2001, krishnamurthiProgrammingLanguagesApplication2007};
however, it also proved to be pedagogically valuable.
I have found the parentheses useful when discussing
scope in conjunction with local-binding features like \lstinline|let|.
This avoids the various confusing ``variable lifting'' semantics found
in languages like Python (\cite{politzPythonFullMonty2013}; see also
posts like \cite{froadieWhatScopeVariable2022}), where the actual defined range of a
variable is not apparent from the source code.

@@

Students are given a working implementation of SMoL, built
as a \lstinline|#lang| language inside Racket \citep{felleisenProgrammableProgrammingLanguage2018}. This language
provides only the defined syntax and semantics of SMoL, with no
(other) Racket features present.
(As in Racket, arguments evaluate left-to-right, to give stateful programs an
unambiguous semantics.) The implementation is available online:
\href{https://github.com/shriram/smol}{https://github.com/shriram/smol}.

@@

There is an extensive body of literature on tutoring systems
(\cite{vanlehnBehaviorTutoringSystems2006a} is a quality survey), and
indeed whole conferences are dedicated to them. I draw on this
literature. In particular, it is common in the literature to talk
about a ``two-loop'' architecture
\citep{vanlehnBehaviorTutoringSystems2006a} where the outer loop
iterates through ``tasks'' (i.e., educational activities) and the
inner loop iterates through UI events within a task. I follow the same
structure in my tutor (\Cref{s:tutor-uiux}).

Many tutoring systems focus on teaching programming (such as the
well-known and heavily studied LISP Tutor
\citep{andersonLISPTutor1985}), and in the process undoubtedly address
some program behavior misconceptions. My tutor differs in a notable
way: it does not try to teach programming per se. Instead, it assumes
a basic programming background and focuses entirely on program
behavior misconceptions and correcting them. I are not aware of a
tutoring system (in computer science) that has this specific design.

@@

The SMoL Tutor is firmly grounded in one technique from
cognitive and educational psychology. The fundamental problem is: how
do you fix a misconception? One approach is to only ``present the
right answer'', for fear that discussing the wrong conception might
actually reinforce it. Instead, there is a body of literature starting
from \cite{posnerTheoryConceptualChange1982} that presents a theory of conceptual
change, at whose heart is the \emph{refutation text}. A refutation
text tackles the misconception directly, discussing and providing a
refutation for the incorrect idea. Several studies \citep{schroederRefutationTextFacilitates2022} have shown
their effectiveness in multiple other domains.

@@

The problem with expert-generated lists is that they can be quite
incomplete. Education researchers have documented the
phenomenon of the \emph{expert blind spot} \citep{nathanExpertBlindSpot2001}: experts
simply do not conceive of many learner difficulties.
Thus, we need
methods to identify problems beyond what experts conceive.

@@

Finally, we are inspired by the significant body of education research
on \emph{concept inventories} \citep{hestenesForceConceptInventory1992} (with a growing
number for computer science, as a survey lists \citep{taylorComputerScienceConcept2014}). In
terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question has
one correct answer and several wrong ones. However, the wrong ones are
chosen with great care. Each one has been validated so that if a
student picks it, we can quite unambiguously determine \emph{what
  misconception the student has}. For instance, if the question is
``What is \lstinline|sqrt(4)|?'', then \lstinline|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \lstinline|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many
  settings. For instance, an educator can use them with clickers to
  get quick feedback from a class. If several students pick a
  specific wrong answer, the educator not only knows they are wrong,
  but also has a strong inkling of \emph{precisely what} misconception
  that group has and can address it directly. I expect our
  instruments to be useful in the same way.}

@@

Our main solution to the expert blind spot is to use the Quizius system
\citep{saarinenHarnessingWisdomClasses2019}. In contrast to the very
heavyweight process (involving a lot of expert time)
that is generally
used to create a concept inventory, Quizius uses a lightweight,
interactive approach to obtain fairly comparable data, which an expert
can then shape into a quality instrument.

@@

Quizius runs interactively during an assignment period.
At each point, it needs to determine which
previously authored program to show a student.
It can
either ``exploit'' a given program that already has responses
or ``explore'' a new
one. Quizius thus treats this as a multi-armed bandit problem
\citep{katehakisMultiArmedBanditProblem1987} and uses that to choose a program.

@@

Prior work using Quizius \citep{saarinenHarnessingWisdomClasses2019} finds that students do author programs
that the experts did not imagine. In our case, we seeded Quizius with
programs from prior papers (\Cref{s:background-misconceptions}), which gives the first
few students programs to respond to. However, we found
that Quizius significantly expanded the scope of our problems and
misconceptions. In our final instrument, most programs were directly
or indirectly inspired by the output of Quizius.

@@

Is our reasoning valid? I have applied standard techniques and
measurements for evaluating student responses. However, our
instruments still lack the validity of a proper concept
inventory. Their creation requires heavyweight processes (such as
Delphi methods \citep{goldmanSettingScopeConcept2010}) that require many hours of expert attention as
well as conversations with learners, and can hence be prohibitive in
cost. The Quizius method (\Cref{s:quizius}) was created precisely to
be an inexpensive method that provides a good proxy.

@@

Our idea of misinterpreters is related to mystery languages
\citep{diwanPLdetectiveSystemTeaching2004,
  pombrioTeachingProgrammingLanguages2017}. Both approaches use
evaluators that represent alternative semantics to the same syntax.
However, the two are complementary.
In mystery languages, instructors design the space of semantics with
pedagogic intent, and
students must create programs to explore that space. Misinterpreters,
in contrast, are driven by student input, while the programs are provided
by instructors. The two approaches also have different goals:
mystery languages focus on encouraging students to experiment with
languages; misinterpreters aim at capturing students' misconceptions.

@@

\bibliography{bibfile}
